---
title: "HDS5101 Final Project"
author: "Klosowski Filip"
date: "2026-03-27"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Introduction

### Research question

How well can health-related variables predict marathon finish time, and which modelling approach performs best?

### Outcome and predictors

- **Outcome:** `actual_finish_time_minutes`
- **Predictors:** `age`, `gender`, `resting_heart_rate_bpm`, `vo2_max`, `bmi`, `sleep_hours_avg`, `injury_count`, `nutrition_score`, `hydration_consistency`

```{r packages}
library(tidyverse)
library(mice)
library(glmnet)
library(mgcv)
library(knitr)

set.seed(21490084)
```

## 2. Data and variable design

```{r data-import}
marathon_data_full <- read_csv("marathondata.csv", show_col_types = FALSE)

analysis_data <- marathon_data_full %>%
  select(
    age,
    gender,
    resting_heart_rate_bpm,
    vo2_max,
    bmi,
    sleep_hours_avg,
    injury_count,
    nutrition_score,
    hydration_consistency,
    actual_finish_time_minutes
  ) %>%
  mutate(gender = as.factor(gender))

dim(analysis_data)
summary(analysis_data$actual_finish_time_minutes)
```

This analysis uses the full marathon dataset as the modelling and evaluation population.

## 3. Exploratory data analysis

```{r missingness-overview}
missingness_tbl <- analysis_data %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "prop_missing") %>%
  arrange(desc(prop_missing))

kable(missingness_tbl, digits = 3)
```

```{r eda-plots, fig.height=7, fig.width=10}
p1 <- ggplot(analysis_data, aes(actual_finish_time_minutes)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.8) +
  labs(title = "Distribution of finish time", x = "Finish time (minutes)", y = "Count")

p2 <- ggplot(analysis_data, aes(gender, actual_finish_time_minutes)) +
  geom_boxplot(fill = "orange", alpha = 0.8) +
  labs(title = "Finish time by gender", x = "Gender", y = "Finish time (minutes)")

p3 <- ggplot(analysis_data, aes(vo2_max, actual_finish_time_minutes)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Finish time vs VO2 max", x = "VO2 max", y = "Finish time (minutes)")

p4 <- ggplot(analysis_data, aes(bmi, actual_finish_time_minutes)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "darkgreen") +
  labs(title = "Finish time vs BMI", x = "BMI", y = "Finish time (minutes)")

p1
p2
p3
p4
```

## 4. Missing data strategies

Three preprocessing strategies are compared:

1. **Complete-case analysis** (`drop_na`)  
2. **Median/mode imputation** (numeric = median, factor = most frequent category)  
3. **Multiple imputation (MICE)** with predictive mean matching

```{r missing-data-prep}
# 1) Complete-case data
complete_case_data <- analysis_data %>%
  drop_na()

# 2) Median/mode imputation helper
mode_value <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}

median_imputed_data <- analysis_data %>%
  mutate(
    across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)),
    across(where(is.factor), ~ forcats::fct_explicit_na(., na_level = as.character(mode_value(.))))
  ) %>%
  mutate(gender = fct_drop(gender))

# 3) MICE imputation
mice_object <- mice(analysis_data, m = 5, method = "pmm", seed = 21490084, print = FALSE)
mice_data <- complete(mice_object, action = 1) %>% mutate(gender = as.factor(gender))

bind_rows(
  tibble(strategy = "Complete-case", n = nrow(complete_case_data)),
  tibble(strategy = "Median/mode", n = nrow(median_imputed_data)),
  tibble(strategy = "MICE", n = nrow(mice_data))
) %>%
  kable()
```

## 5. Modelling approach (full population evaluation)

Models compared within each preprocessing strategy:

- **Linear regression** (baseline)
- **LASSO regression** (`glmnet` with CV for lambda)
- **GAM** (smooth terms for continuous predictors)

```{r model-functions}
model_formula_lm <- actual_finish_time_minutes ~ age + gender + resting_heart_rate_bpm +
  vo2_max + bmi + sleep_hours_avg + injury_count + nutrition_score + hydration_consistency

model_formula_gam <- actual_finish_time_minutes ~ gender +
  s(age) + s(resting_heart_rate_bpm) + s(vo2_max) + s(bmi) +
  s(sleep_hours_avg) + s(injury_count) + s(nutrition_score) + s(hydration_consistency)

get_metrics <- function(actual, predicted) {
  tibble(
    RMSE = sqrt(mean((actual - predicted)^2)),
    MAE = mean(abs(actual - predicted)),
    R_squared = cor(actual, predicted)^2
  )
}

fit_and_score <- function(data, strategy_name) {
  # Linear model
  lm_fit <- lm(model_formula_lm, data = data)
  lm_pred <- predict(lm_fit, newdata = data)
  lm_metrics <- get_metrics(data$actual_finish_time_minutes, lm_pred) %>%
    mutate(model = "Linear regression", preprocessing = strategy_name)

  # LASSO model
  x <- model.matrix(model_formula_lm, data = data)[, -1]
  y <- data$actual_finish_time_minutes
  cvfit <- cv.glmnet(x, y, alpha = 1, nfolds = 10)
  lasso_pred <- as.numeric(predict(cvfit, s = "lambda.min", newx = x))
  lasso_metrics <- get_metrics(y, lasso_pred) %>%
    mutate(model = "LASSO", preprocessing = strategy_name)

  # GAM model
  gam_fit <- gam(model_formula_gam, data = data, method = "REML")
  gam_pred <- predict(gam_fit, newdata = data)
  gam_metrics <- get_metrics(data$actual_finish_time_minutes, gam_pred) %>%
    mutate(model = "GAM", preprocessing = strategy_name)

  bind_rows(lm_metrics, lasso_metrics, gam_metrics)
}
```

```{r run-models}
results_tbl <- bind_rows(
  fit_and_score(complete_case_data, "Complete-case"),
  fit_and_score(median_imputed_data, "Median/mode imputation"),
  fit_and_score(mice_data, "MICE")
) %>%
  select(model, preprocessing, RMSE, MAE, R_squared) %>%
  arrange(preprocessing, RMSE)

kable(results_tbl, digits = 3)
```

## 6. Results and interpretation

- The table above provides a direct comparison of modelling performance using the **full dataset as the population for evaluation**.
- Linear regression provides a clear baseline.
- LASSO highlights whether coefficient shrinkage helps predictive stability.
- GAM shows whether nonlinear patterns improve predictive fit.

```{r lasso-coef-example}
# Example: variable importance via non-zero coefficients for MICE + LASSO
x_mice <- model.matrix(model_formula_lm, data = mice_data)[, -1]
y_mice <- mice_data$actual_finish_time_minutes
cvfit_mice <- cv.glmnet(x_mice, y_mice, alpha = 1, nfolds = 10)
coef_tbl <- as.matrix(coef(cvfit_mice, s = "lambda.min")) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  rename(coefficient = `1`) %>%
  filter(coefficient != 0)

kable(coef_tbl, digits = 3)
```

## 7. Discussion

This workflow was structured to stay focused and comparable:

1. One research question (prediction).
2. One consistent set of predictors.
3. Three explicit missing-data strategies.
4. Three model families compared with identical metrics.
5. Full-population evaluation to match the project requirement.

## 8. Conclusion

Health-related variables provide measurable predictive value for marathon finish time, with model performance varying by both preprocessing strategy and model class. Comparing complete-case, median/mode imputation, and MICE alongside linear regression, LASSO, and GAM provides a stronger and more deliberate workflow for reporting final results.
