---
title: "HDS5101_final_vid"
author: "Klosowski Filip"
date: "2026-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)

library(mice)
library(Hmisc)
library(VIM)

set.seed(21490084)
```

# Question

To what extent do health-related variables influence marathon finish time, and how well can these variables be used to predict performance?

# Full Data Set Exploration

```{r}
marathon_data_full <- read_csv("marathondata.csv")
head(marathon_data_full, 10)
dim(marathon_data_full)
aggr(marathon_data_full)

missing_prop_long <- marathon_data_full %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "prop_missing") %>%
  filter(prop_missing > 0) %>%
  
  arrange(desc(prop_missing))

missing_prop_long
```

n = 80,000

p = 41

Study type = Observational cohort study

Population = Marathon runners from a fictional running club

injury_severity is structurally missing when injury_count = 0 so I will remove it. runner_id and marathon_date seem redundant. Weekly_mileage_miles is a duplicate variable of weekly_mileage_km. gender, training_program, marathon_weather, course_difficulty need to be converted into factors.

I will sub sample the data into 2,000 observations for computational efficiency and to make the data set more realistic. I will also split it into a training and testing sets.

# Sample Data

```{r}
#sample 2,000 observations

marathon_data_sampled <- marathon_data_full %>%
  slice_sample(n = 2000)

```

# Data wraggling

```{r}

marathon_data_sampled <- marathon_data_sampled %>%
  select(
    -runner_id,
    -injury_severity,
    -weekly_mileage_miles,
    -marathon_date #redundant
  ) %>%
  mutate(
    gender = factor(gender),
    training_program = factor(training_program),
    marathon_weather = factor(marathon_weather),
    course_difficulty = factor(course_difficulty) #convert to factor
  )
```



```{r}
head(marathon_data_sampled, 10)
dim(marathon_data_sampled)
aggr(marathon_data_sampled)

missing_prop_long2 <- marathon_data_sampled %>%
  summarise(across(everything(), ~ mean(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "prop_missing") %>%
  filter(prop_missing > 0) %>%
  arrange(desc(prop_missing))

missing_prop_long2


```

n = 2,000

p = 37

The observations with missing outcome variables will be dropped, then I will compare complete case analysis (CC) to median imputation to multiple imputation for missing predictors (MICE). To evaluate which is the best I will run a simple LM first.

```{r}
marathon_data_sampled <- marathon_data_sampled %>%
  drop_na(actual_finish_time_minutes)
```


# Split Data

```{r}
set.seed(21490084)

train_index <- sample(
  seq_len(nrow(marathon_data_sampled)),
  size = 0.7 * nrow(marathon_data_sampled)
)

train_data <- marathon_data_sampled[train_index, ]
test_data  <- marathon_data_sampled[-train_index, ]

dim(train_data)
dim(test_data)

```

# Missing Data

**Complete Case**

```{r}
train_cc <- train_data %>%
  drop_na()

test_cc <- test_data %>%
  drop_na()
```

**Median Imputation**

```{r}

median_impute <- function(x) {
  if (is.numeric(x)) {
    x[is.na(x)] <- median(x, na.rm = TRUE)
  }
  x
}


train_median <- train_data %>%
  mutate(across(-actual_finish_time_minutes, median_impute))

test_median <- test_data %>%
  mutate(across(-actual_finish_time_minutes, median_impute))
```


The observations with missing outcome variables will be dropped, then I will compare complete case analysis to median imputation to multiple imputation for missing predictors (MICE). To evaluate which is the best I will run a simple LM first.

**MICE**







